---
title: "Ellerbe_Analysis"
author: "Analise Lindborg"
date: "3/22/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_knit$set(root.dir = '/Users/analiselindborg/Desktop/Desktop - Analise’s MacBook Pro/Data Analytics/EDA_Final_Project_DurhamWQ')
```

## Setting up R session

```{r, message = FALSE}
getwd()
library(tidyverse)
library(lubridate)

ellerbe.data <- read.csv("./Data/Raw/Ellerbe Creek/durham_AllData.csv")
```

## Data cleanup


```{r}
## 1. Select columns of interest
# I could not find the metadata. This may be important for the QC column (I think U in the QC column means that the value was set to the detection limit (because the value was likely below the detection limit), but I am not sure). Also not sure what flow severity is or if we would want to use it. 
ellerbe.sub <- ellerbe.data %>%
  select(Stream.Name, Station.Name, Parameter, Date.Time, Value, QA.Code, Unit, Comments, Rain.in.Last.24.Hours, Sky.Condition)

## 2. Subsetting data for following metrics: DO, pH, Fecal Coliform, Zn, Cu, temp, TP, TSS, and turbidity and sites 1.9 and 7.1 (5.6 does not have 2020 data)

ellerbe.sub <- ellerbe.sub %>%
  filter(Parameter %in% c("Dissolved Oxygen", "pH", "Fecal Coliform", "Zinc", "Copper", "Temperature", "Total Phosphorus", "Total Suspended Solids", "Turbidity")) %>%
  filter(Station.Name != "EL5.6EC")

## 3. Fix date
ellerbe.sub <- ellerbe.sub %>%
  separate(Date.Time, into = c("Date", "Time"), sep = "\\s")

ellerbe.sub$Date <- as.Date(ellerbe.sub$Date, format = "%m/%d/%y")

## 4. Some sample dates have site duplicates. Averaging these duplicate values to get one value for each parameter at each site on each data. Create a year column
ellerbe.grouped <- ellerbe.sub %>%
  group_by(Stream.Name, Station.Name, Parameter, Date, Comments, Rain.in.Last.24.Hours, Sky.Condition) %>%
 summarize(Final_Value = mean(Value)) %>%
  mutate(Year = year(Date))

ellerbe.grouped$Year <- as.character(ellerbe.grouped$Year)

## save final dataset
#write.csv(ellerbe.all, "./Data/Processed/Ellerbe_processed.csv")
```

Analysis ideas
  # look at other parameters individually and compare between years, keeping sites separate
  # Combine Zn and copper?
  # look at temp and DO and temp and fecal coliform
    # bacteria is temp and flow driven
  # tss and turbidity comparison (as indicator of which to do and inputs) - scatterplot
  # temp and sky condition (as indicator of cover) - box plot
  # spikes in turbid with rain in last 24 hours - box plot

```{r}
theme <- 
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        text = element_text(color = "black", size = 10),
        axis.text.x = element_text(color = "black"),
        axis.text.y = element_text(color = "black"))

theme_set(theme)
```
Visualizing changes in all parameters over time

DO: There are no distinguishable trends over time and do not appear to be differences between the two sites. Looks like there may be seasonality, we could use seasonal Mann Kendall to explore further

pH: Looks like there may be a downward trend from 2019 to 2020. Run lm stats on this. 

Fecal Coliform: Relatively low with spikes at both stations in August of both years. Could be indicative of seasonal variations in natural E. coli (or other bacteria)? I am not sure what they culture for in this study. Seems unlikely this is gross wastewater effluent contamination since it is seasonal. 

Temperature: Temp measurements are sporadic (there are gaps in data) and temp is seasonal (obviously). Summer and winter temps appear similar for 2019 and 2020, but hard to tell beccause of data gaps. Could do some extrapolation?

TP: Could be a slight increase over time, site 1.9 also has higher spikes. Appears to be some seasonality but difficult to tell

TSS: Again, could be slight increase over time and site 1.9 has spike fall

turbidity: Appears similar to TSS
```{r}
#creating pivot wider dataframe to make WQ parameters their own columns. 
ellerbe.wide <- ellerbe.grouped %>%
  pivot_wider(names_from = Parameter, values_from = Final_Value)

#DO
ggplot(ellerbe.wide, aes(x = Date, y = `Dissolved Oxygen`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Dissolved Oxygen (mg/L)", x = "Date") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#pH
ggplot(ellerbe.wide, aes(x = Date, y = pH, color = Station.Name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "pH", x = "Date") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Fecal Coliform
ggplot(ellerbe.wide, aes(x = Date, y = `Fecal Coliform`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Fecal Coliform (cfu/100mL)", x = "Date") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#Temp
ggplot(ellerbe.wide, aes(x = Date, y = Temperature, color = Station.Name)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Temperature (Celcius)", x = "Date") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#TP
ggplot(ellerbe.wide, aes(x = Date, y = `Total Phosphorus`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Total Phosphorus (mg/L)", x = "Date") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#TSS
ggplot(ellerbe.wide, aes(x = Date, y = `Total Suspended Solids`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Total Suspended Solids (mg/L)", x = "Date") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#turbidity
ggplot(ellerbe.wide, aes(x = Date, y = `Turbidity`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Turbidity (NTU)", x = "Date") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Explore metal concentrations at each of the sites between 2019 and 2020

```{r}
ggplot(subset(ellerbe.grouped, Parameter %in% c("Zinc", "Copper")), aes(x = Date, y = Final_Value, color = Parameter)) +
  geom_point() +
  scale_color_viridis_d(option = "magma", begin = 0.8, end = 0.2, name = "Metal") +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  labs(y = "Concentration (ug/L)", x = "Date") +
  theme+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Station.Name, nrow = 2)
```


Comparison of temperature and sky conditions. May indicate canopy cover (tell us something about the sites and riparian area). 

Does not appear that there is any relationship here. I think we would need more finite data (like hourly) to make this kind of determination. 
```{r}
ggplot(subset(ellerbe.grouped, Parameter == "Temperature"), aes(x = Sky.Condition, y = Final_Value, fill = Station.Name)) +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Temperature (Celcius)", x = "Sky Condition") +
  theme+
  theme(panel.grid.major = element_blank())

#Over time (since this likely changes with season)
ggplot(subset(ellerbe.grouped, Parameter == "Temperature"), aes(x = Date, y = Final_Value, color = Sky.Condition)) +
  geom_point() +
  scale_color_viridis_d(option = "magma", begin = 0.8, end = 0.2, name = "Sky Condition") +
  scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  labs(y = "Temperature (Celcius)", x = "Date") +
  theme+
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Station.Name, nrow = 2)
```

Comparison of turbidity and rain. Determine if the sites are highly influenced by storm events. Could also be an indication of erosion. This could fit with general exploration of the site and trying to set up context for this being a highly degraded creek in need of restoration/additional monitoring. Could compare this to observations of Eno River sites. Comparison condudcted for 2019 and 2020 combined Not sure it totally fits with project scope/year comparisons...

Figure shows that turbidity is higher after rain event, indicating sediment loading.
```{r}
ggplot(subset(ellerbe.grouped, Parameter == "Turbidity"), aes(x = Rain.in.Last.24.Hours, y = Final_Value, fill = Station.Name)) +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Turbidity (NTU)", x = "Rain in Last 24 Hours?") +
  theme+
  theme(panel.grid.major = element_blank())

#without outlier (this changed how the data calucated the boxplots, not sure how to fix that...)
ggplot(subset(ellerbe.grouped, Parameter == "Turbidity"), aes(x = Rain.in.Last.24.Hours, y = Final_Value, fill = Station.Name)) +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  ylim(0, 100) +
  labs(y = "Turbidity (NTU)", x = "Rain in Last 24 Hours?") +
  theme +
  theme(panel.grid.major = element_blank())
```

Comparison of turbidity and TSS. How well do these correlate? Indicative of certain types of pollution

Measurements appear highly correlated. Need to add statistics. But this could be indicative of certain types of pollution such as sediment, algae, etc. This also indicates that Durham could rely on turbidity alone as a measure for clarity. TSS is expensive to sample.
```{r}
ggplot(ellerbe.wide, aes(x = Turbidity, y = `Total Suspended Solids`, color = Station.Name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Total Suspended Solids (mg/L)", x = "Turbidity (NTU)") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#removing outlier point to better show trend
ggplot(ellerbe.wide, aes(x = Turbidity, y = `Total Suspended Solids`, color = Station.Name)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlim(0,80)+
  ylim(0,80) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Total Suspended Solids (mg/L)", x = "Turbidity (NTU)") +
  theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```