---
title: "Eno_Analysis"
author: "Olivia August"
date: "4/5/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r}
#get directory
getwd()

#load packages
library(tidyverse)
library(lubridate)
library(sf)
#remotes::install_github("r-spatial/mapview")
library(mapview)
library(ggplot2)
```

```{r}
# import all data
Durham_stations <- read.csv('../Data/Raw/durham_station.csv',header = TRUE, row.names=NULL)[,1:3]
colNames <- read.csv('../Data/Raw/Eno_River/durham_data_Cu.csv',skip = 11, nrows=1, header = F)
Eno_River_Cu <- read.csv('../Data/Raw/Eno_River/durham_data_Cu.csv',skip = 12, row.names=NULL, header = F)[,1:17]
Eno_River_DO <- read.csv('../Data/Raw/Eno_River/durham_data_DO.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_FecCol <- read.csv('../Data/Raw/Eno_River/durham_data_FecCol.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_pH <- read.csv('../Data/Raw/Eno_River/durham_data_pH.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_Temp <- read.csv('../Data/Raw/Eno_River/durham_data_Temp.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_TP <- read.csv('../Data/Raw/Eno_River/durham_data_TP.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_TSS <- read.csv('../Data/Raw/Eno_River/durham_data_TSS.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_Turb <- read.csv('../Data/Raw/Eno_River/durham_data_Turb.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_Zn <- read.csv('../Data/Raw/Eno_River/durham_data_Zn.csv',skip = 12, header = F, row.names=NULL)[,1:17]

# compile into one dataframe
Eno_River_All_raw <- rbind(Eno_River_Cu,Eno_River_DO,Eno_River_FecCol,Eno_River_pH,Eno_River_Temp,Eno_River_TP,Eno_River_TSS,Eno_River_Turb,Eno_River_Zn)

# name columns
colnames(Eno_River_All_raw) <- c("Stream.Name","Station.Name",colNames[3:4],"Date.Time",colNames[6],"QA.Code",colNames[8:10],"Rain.in.Last.24.Hours","Sky.Condition",colNames[13:17])
```


```{r}
# keep relevant columns for analysis
Eno_River_parameters <- Eno_River_All_raw%>% 
  select(Stream.Name, Station.Name, Parameter, Date.Time, Value, QA.Code, Unit, Comments, Rain.in.Last.24.Hours, Sky.Condition) 

# Make date correct form
Eno_River_parameters<- Eno_River_parameters %>%
  separate(Date.Time, into = c("Date", "Time"), sep = "\\s")
Eno_River_parameters$Date <- as.Date(Eno_River_parameters$Date, format = "%m/%d/%Y")

# average duplicates 
Eno_River_parameters_clean <- Eno_River_parameters %>%
  group_by(Stream.Name, Station.Name, Parameter, Date, Comments, Rain.in.Last.24.Hours, Sky.Condition) %>%
  summarize(Final_Value = mean(Value)) %>%
  mutate(Month = month(Date))%>%
  mutate(Year = year(Date))

#split parameters into their own columns
Eno_River_parameters_split <- Eno_River_parameters_clean %>%
  pivot_wider(names_from = Parameter, values_from = Final_Value)


#join with Station coordinate data
Eno_River_parameters_stations <- merge(x = Eno_River_parameters_split,
                                       y=Durham_stations,
                                       by.x = "Station.Name",
                                       by.y = "Name")

#split into 2019 and 2020
Eno_River_Data_2019 <- Eno_River_parameters_stations %>% 
  filter(between(Date,as.Date("2019-01-01"),as.Date("2019-12-31")))
Eno_River_Data_2020 <- Eno_River_parameters_stations %>% 
  filter(between(Date,as.Date("2020-01-01"),as.Date("2020-12-31")))

## save processed data
#write_csv(Eno_River_parameters_stations, '../Data/Processed/Eno_processed.csv')

#create spatial dataframe
Eno_River_parameters_stations_sf <- Eno_River_parameters_stations %>% 
  st_as_sf(coords = c('Longitude','Latitude'),crs=4269)
mapview(Eno_River_parameters_stations_sf)
```


```{r}

theme <- 
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        text = element_text(color = "black", size = 10),
        axis.text.x = element_text(color = "black"),
        axis.text.y = element_text(color = "black"))

theme_set(theme)
```


```{r}

#DO
DO_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Dissolved Oxygen`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Dissolved Oxygen (mg/L)", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(DO_Year.Plot)
  
#ph 
pH_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = pH, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "pH", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(pH_Year.Plot)

#Fecal Coliform 
FC_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Fecal Coliform`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  scale_y_log10()+
  labs(y = "Fecal Coliform (cfu/100mL)", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(FC_Year.Plot)

#Temperature
Temp_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = Temperature, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Temperature (C)", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(Temp_Year.Plot)

#TP 
TP_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Total Phosphorus`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Total Phosphorus (mg/L)", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(TP_Year.Plot)

#TSS
TSS_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Total Suspended Solids`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Total Suspended Solids (mg/L)", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(TSS_Year.Plot)

#Turbidity
Turbidity_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = Turbidity, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Turbidity (NTU)", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(Turbidity_Year.Plot)


#Metals 
Metals_Year.Plot <- ggplot(subset(Eno_River_parameters_clean,Parameter %in% c("Zinc", "Copper")),aes(x = Month, y = Final_Value, color = Parameter, shape = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(Year)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Metal Concentration (ug/L)", x = "Date") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
show(Metals_Year.Plot)
```

Summarize Results 
DO: The data for Dissolved Oxygen has a simlar trend across stations and across years. The smoothed plot line shows that Dissolved Oxygen was overall slightly higher in 2019 than in 2020.

pH: The smoothed line for pH is similar across stations and from 2019 to 2020. However, the maximum pH during the Summer peak is high in 2019, ~ 7.5, than in 2020, 7.


```{r}
#ANOVA for pH
# *still need to run this
aov <- aov(Eno_River_parameters_split$pH ~ Eno_River_parameters_split$Year)
summary(aov)
TukeyHSD(aov)
```

Fecal Coliform: The fecal coliform has no discernible trend in both the 2019 and 2020 grpahs. However, the smoothed line shows an overall decrease in Fecal Coliform concentration acrosss the year for 2019. However, the 2020 data shows an overall increase in Fecal Coliform across the year. The max values in 2020 are much higher, > 10,000, than in 2019, < 10,000.

Temperature: The temperature data across stations and years showsa similar pattern of max temperatures occuring in the late summer months. There is no significant difference in temperature between 2019 and 2020.

TP: Total phoshorus has no discenible pattern in the 2019 or 2020 plots. The concentration increases across the year in 2020 and decreases in 2020 based on the smoothed graph. Overall, concentrations seem relatively comparable besides a greater max value in the 2019 data.

TSS: The TSS shows very different patterns across 2019 and 2020. There is a peas in Sring of 2019 and late Summer in 2020. The smoothed line shows that overall greater concentrations of TSS were observed in 2020.

Turb: The plots of Turbidity show very different trends across 2019 and 2020. There is a peak in the Spring of 2019 and the late Summer of 2020. The overall concentrations of Turbidity are greater in 2020.

Metals: The concentrations of Zinc are much higher than Copper in both 2019 and 2020 plots. The Zinc concentrations are much higher in 2019 than in 2020. The Copper concentrations do not vary significantly between 2019 and 2020. 

```{r}
##ANOVA for metals
aov <- aov(Eno_River_parameters_split$Zinc ~ Eno_River_parameters_split$Year)
summary(aov)
 
aov <- aov(Eno_River_parameters_split$Copper ~ Eno_River_parameters_split$Year)
summary(aov)  
```

```{r}
#Temp and Sky condition
Temp_Sky.plot <- ggplot(subset(Eno_River_parameters_clean, Parameter == "Temperature"), aes(x = Sky.Condition, y = Final_Value, fill = Station.Name)) +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Temperature (Celcius)", x = "Sky Condition") +
  theme+
  theme(panel.grid.major = element_blank())
show(Temp_Sky.plot)

#Temp and Sky condition over time
Temp_Sky_ts.plot <- ggplot(subset(Eno_River_parameters_clean, Parameter == "Temperature"), aes(x = Month, y = Final_Value, color = Sky.Condition, shape = Station.Name)) +
  geom_point() +
  scale_color_viridis_d(option = "magma", begin = 0.8, end = 0.2, name = "Sky Condition") +
  #scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  labs(y = "Temperature (Celcius)", x = "Date") +
  theme+
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Year, nrow = 2)
show(Temp_Sky_ts.plot)
```

There is not significant difference in measured temperature based on the observed Sky Condition. The variation across the various Sky Conditions is not consistent across the different monitoring stations. However, there is a noticeable difference in temperature across stations which may give information about the monitoring location.

Across the year, the there is not noticeable trned between the Sky Condition and the temperature. The temperature seems to be more seasonally affected than by the Sky Condition.


```{r}
#turbidity and Recent Rain
Turbidity_Rain.plot <- ggplot(subset(Eno_River_parameters_clean, Parameter == "Turbidity"), aes(x = Rain.in.Last.24.Hours, y = Final_Value, fill = Station.Name)) +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Turbidity (NTU)", x = "Rain in Last 24 Hours?") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow = 2)
show(Turbidity_Rain.plot)


##ANOVA

#comparing turbidity and rain in last 24 hours for 2019 (all sites combined, combined sites because we have different numnber of sites between 2019 and 2020 so want to make the comparison similar)
d.2019 <- Eno_River_parameters_split %>%
  filter(Year == 2019)

aov <- aov(d.2019$Turbidity ~ d.2019$Rain.in.Last.24.Hours)
summary(aov)

#comparing turbidity and rain in last 24 hours for 2020
d.2020 <- Eno_River_parameters_split %>%
  filter(Year == 2020)

aov <- aov(d.2020$Turbidity ~ d.2020$Rain.in.Last.24.Hours)
summary(aov)


#compare turbidity between 2019 and 2020 
aov <- aov(Eno_River_parameters_split$Turbidity ~ Eno_River_parameters_split$Year)
summary(aov)
TukeyHSD(aov)
```

There is noticeable difference in the Turbidity and recent rain. This trend is less noticeable in the 2019, but the 2020 data shows a clear disparity between recent rain and the turbidity readings.

```{r}
#TSS & Turbidity
Turbidity_TSS.plot <- ggplot(subset(Eno_River_parameters_clean,Parameter %in% c("Turbidity", "Total Suspended Solids")), aes(x = Month, y = Final_Value, color = Parameter, shape= Station.Name)) +
  geom_point() +
  geom_line()+
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Turbidity (NTU)", x = "Rain in Last 24 Hours?") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow = 2)
show(Turbidity_TSS.plot)

```

The plot shows that TSS and Turbidity readings are closely linked and follow the same trend across stations and across years. These parameters have similar peaks. There is some variation across stations, but the pattern remains the same across.


```{r}
# Temp & DO 
Temp_DO.plot <- ggplot(Eno_River_parameters_stations, aes(x = Temperature, y = `Dissolved Oxygen`, color = Station.Name)) +
  geom_point() +
  geom_line()+
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Dissolved Oxygen", x = "Temperature (Celcius)") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow = 2)
show(Temp_DO.plot)


#Temp & Fec Col
Temp_FC.plot <- ggplot(Eno_River_parameters_stations, aes(x = Temperature, y = `Fecal Coliform`, color = Station.Name)) +
  geom_point() +
  geom_line()+
  scale_fill_viridis_d(begin = 0.6, end = 0.3, name = "Sampling Station") +
  labs(y = "Fecal Coliform (cfu/100mL)", x = "Temperature (Celcius)") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow = 2)
show(Temp_FC.plot)
```

##DO & Temp
There is a clear relationship with Temperature and DO. The DO decreases linerarly as temperature increases. This is true for all monitoring stations and both years.

##FC & Temp
For the 2019 data there is not a clear relationship between Fecal Coliform concentration and temperature. However, in 2020 there is a peak in the Fecal Coliform concentration at the higher temperatures. It can not be concluded that there is a relationship between these parameters, but further investigation should occur. 
