---
title: "Eno_Analysis"
author: "Olivia August"
date: "4/5/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r echo=FALSE}
#get directory
getwd()

#load packages
library(tidyverse)
library(lubridate)
library(sf)
#remotes::install_github("r-spatial/mapview")
library(mapview)
library(ggplot2)
library(agricolae)
```

```{r echo=FALSE}
# import all data
Durham_stations <- read.csv('../Data/Raw/durham_station.csv',header = TRUE, row.names=NULL)[,1:3]
colNames <- read.csv('../Data/Raw/Eno_River/durham_data_Cu.csv',skip = 11, nrows=1, header = F)
Eno_River_Cu <- read.csv('../Data/Raw/Eno_River/durham_data_Cu.csv',skip = 12, row.names=NULL, header = F)[,1:17]
Eno_River_DO <- read.csv('../Data/Raw/Eno_River/durham_data_DO.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_FecCol <- read.csv('../Data/Raw/Eno_River/durham_data_FecCol.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_pH <- read.csv('../Data/Raw/Eno_River/durham_data_pH.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_Temp <- read.csv('../Data/Raw/Eno_River/durham_data_Temp.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_TP <- read.csv('../Data/Raw/Eno_River/durham_data_TP.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_TSS <- read.csv('../Data/Raw/Eno_River/durham_data_TSS.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_Turb <- read.csv('../Data/Raw/Eno_River/durham_data_Turb.csv',skip = 12, header = F, row.names=NULL)[,1:17]
Eno_River_Zn <- read.csv('../Data/Raw/Eno_River/durham_data_Zn.csv',skip = 12, header = F, row.names=NULL)[,1:17]

# compile into one dataframe
Eno_River_All_raw <- rbind(Eno_River_Cu,Eno_River_DO,Eno_River_FecCol,Eno_River_pH,Eno_River_Temp,Eno_River_Turb,Eno_River_Zn)

# name columns
colnames(Eno_River_All_raw) <- c("Stream.Name","Station.Name",colNames[3:4],"Date.Time",colNames[6],"QA.Code",colNames[8:10],"Rain.in.Last.24.Hours","Sky.Condition",colNames[13:17])
```


```{r echo=FALSE}
# Make date correct form
Eno_River_parameters_sub<- Eno_River_All_raw %>%
  separate(Date.Time, into = c("Date", "Time"), sep = "\\s")
Eno_River_parameters_sub$Date <- as.Date(Eno_River_parameters_sub$Date, format = "%Y-%m-%d")

# fix date for TSS and TP
Eno_River_TSS_TP <- rbind(Eno_River_TP,Eno_River_TSS)
colnames(Eno_River_TSS_TP) <- c("Stream.Name","Station.Name",colNames[3:4],"Date.Time",colNames[6],"QA.Code",colNames[8:10],"Rain.in.Last.24.Hours","Sky.Condition",colNames[13:17])
Eno_River_TSS_TP<- Eno_River_TSS_TP %>%
  separate(Date.Time, into = c("Date", "Time"), sep = "\\s")
Eno_River_TSS_TP$Date <- as.Date(Eno_River_TSS_TP$Date, format = "%m/%d/%Y")

#combine TSS and TP with rest
Eno_River_parameters<- rbind(Eno_River_parameters_sub,Eno_River_TSS_TP)

# keep relevant columns for analysis
Eno_River_parameters <- Eno_River_parameters%>% 
  select(Stream.Name, Station.Name, Parameter, Date, Value, QA.Code, Unit, Comments, Rain.in.Last.24.Hours, Sky.Condition) 



# average duplicates 
Eno_River_parameters_clean <- Eno_River_parameters %>%
  group_by(Stream.Name, Station.Name, Parameter, Date, Comments, Rain.in.Last.24.Hours, Sky.Condition) %>%
  summarize(Final_Value = mean(Value)) %>%
  mutate(Month = month(Date))%>%
  mutate(Year = year(Date))%>%
  filter(Station.Name %in% c("EN13.3ER", "EN4.9ER", "EN8.9ER" ))

#split parameters into their own columns
Eno_River_parameters_split <- Eno_River_parameters_clean %>%
  pivot_wider(names_from = Parameter, values_from = Final_Value) %>%
  drop_na(Date)

#join with Station coordinate data
Eno_River_parameters_stations <- merge(x = Eno_River_parameters_split,
                                       y=Durham_stations,
                                       by.x = "Station.Name",
                                       by.y = "Name")

#split into 2019 and 2020
Eno_River_Data_2019 <- Eno_River_parameters_stations %>% 
  filter(between(Date,as.Date("2019-01-01"),as.Date("2019-12-31")))
Eno_River_Data_2020 <- Eno_River_parameters_stations %>% 
  filter(between(Date,as.Date("2020-01-01"),as.Date("2020-12-31")))

## save processed data
#write_csv(Eno_River_parameters_clean, '../Data/Processed/Eno_processed.csv')
#write_csv(Eno_River_parameters_stations, '../Data/Processed/Eno_processed_wide.csv')

#create spatial dataframe
Eno_River_parameters_stations_sf <- Eno_River_parameters_stations %>% 
  st_as_sf(coords = c('Longitude','Latitude'),crs=4269)
mapview(Eno_River_parameters_stations_sf)
```


```{r echo=FALSE}

theme <- 
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        text = element_text(color = "black", size = 10),
        axis.text.x = element_text(color = "black"),
        axis.text.y = element_text(color = "black"))

theme_set(theme)
```


```{r echo=FALSE}
#DO
DO_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Dissolved Oxygen`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 4, lty = 2) +
  geom_text(x = 12, y = 4.5, aes(label = "Lower Limit"), hjust = 1) +
  facet_wrap(~Year, nrow=2) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  labs(y = "Dissolved Oxygen (mg/L)", x = "Month") +
  theme(axis.text.x = element_text(hjust = 1))+
  theme_bw()
show(DO_Year.Plot)
  
#ph 
pH_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = pH, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_wrap(~Year, nrow=2) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_hline(yintercept = 6, lty = 2) +
  geom_hline(yintercept = 9, lty = 2, color = "red") +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  labs(y = "pH", x = "Month") +
  theme_bw()
show(pH_Year.Plot)

#Fecal Coliform 
FC_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Fecal Coliform`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 400, lty = 2, color = "red") +
  scale_y_log10(
    breaks = scales::trans_breaks("log10", function(x) 10^x),
    labels = scales::trans_format("log10", scales::math_format(10^.x))) +
  annotation_logticks(sides = "l", scaled = TRUE) +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  facet_wrap(~Year, nrow=2) +
  labs(y = "Fecal Coliform (cfu/100mL)", x = "Month") +
  theme 
show(FC_Year.Plot)

#Temperature
Temp_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = Temperature, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_wrap(~Year, nrow=2) +
  geom_hline(yintercept = 32, lty = 2) +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  labs(y = "Temperature (C)", x = "Month") +
  theme(axis.text.x = element_text(hjust = 1))+
  theme_bw()
show(Temp_Year.Plot)

#TP 
TP_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Total Phosphorus`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_wrap(~Year, nrow=2) +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  labs(y = "Total Phosphorus (mg/L)", x = "Month") +
  theme(axis.text.x = element_text(hjust = 1))+
  theme_bw()
show(TP_Year.Plot)

#TSS
TSS_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = `Total Suspended Solids`, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_wrap(~Year, nrow=2) +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  labs(y = "Total Suspended Solids (mg/L)", x = "Month") +
  theme(axis.text.x = element_text(hjust = 1))+
  theme_bw()
show(TSS_Year.Plot)

#Turbidity
Turbidity_Year.Plot <- ggplot(Eno_River_parameters_stations, aes(x = Month, y = Turbidity, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_wrap(~Year, nrow=2) +
  geom_hline(yintercept = 50, lty = 2) +
  geom_text(x = 12, y = 58, aes(label = "Upper Limit"), hjust = 1) +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  labs(y = "Turbidity (NTU)", x = "Month") +
  theme(axis.text.x = element_text(hjust = 1))+
  theme_bw()
show(Turbidity_Year.Plot)
```

As part of the exploratory analysis, each water quality parameter from the monitoring data at the Eno River Stations was visualized by year for 2019 and 2020. The following section includes the plots and descriptions of the visualized water quality data by year. 

DO: The data for Dissolved Oxygen has a similar trend across stations and across years. The smoothed plot line shows that Dissolved Oxygen was overall slightly higher in 2019 than in 2020.

pH: The smoothed trend line for pH is similar across stations and from 2019 to 2020. However, the maximum pH during the Summer peak is high in 2019, ~ 7.5, than in 2020, 7. The pH data for both years shows that the pH in the Eno River is within in the range permissible by North Carolina Water Quality Parameters of 6-9.

Fecal Coliform: The Fecal Coliform has no discernible seasonal trend in both the 2019 and 2020 graphs. Most of the measurements for both years fall between 100 and 10,000. However, the maximum Fecal Coliform concentrations in 2020 are much higher than 2019. The maximum measured concentration in 2019 was approximately 5,000 and the maximum measured concentration in 2020 was around 35,000.

Temperature: The temperature data across stations and years shows a similar pattern of max temperatures occurring in the late summer months. There appears to be no significant difference in temperature between 2019 and 2020.

TP: Total phosphorus has no discernible pattern in the 2019 or 2020 plots. Overall, concentrations seem relatively comparable besides a greater maximum measured value in the 2019 data.

TSS: The TSS shows very different patterns across 2019 and 2020. There are peaks in concentration in the Spring of 2019 and the late Summer in 2020. Based on the plots, there appears to be greater TSS measured in 2020 than 2019.

Turbidity: The plots of turbidity show very different trends across 2019 and 2020. Similar to TSS, there is a peak in the Spring of 2019 and the late Summer of 2020. Based on the plots, the concentrations of Turbidity seem to be higher in 2020.


```{r echo=FALSE}
#pH Significance 
pH.regression <- lm(Eno_River_parameters_stations$pH ~ Eno_River_parameters_stations$Year)
summary(pH.regression)

```
Based on the graphs in the exploratory analysis the pH readings appear to be similar in 2019 and 2020. This was confirmed by performing a linear regression which concluded that the difference in pH from 2019 to 2020 is not statistically significant.


```{r echo=FALSE}
#Metals 
Metals_Year.Plot <- ggplot(subset(Eno_River_parameters_clean,Parameter %in% c("Zinc", "Copper")),aes(x = Month, y = Final_Value, color = Station.Name)) +
  geom_point() +
  geom_line() +
  facet_grid(Year~Parameter) +
  scale_color_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  scale_x_continuous(breaks=seq(1,12,1)) +
  labs(y = "Concentration (ug/L)", x = "Month") +
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 1))

show(Metals_Year.Plot)
```

Metals: Comparing the concentrations of Copper The Zinc concentrations are much higher in 2019 than in 2020. This is expected based on the North Carolina Water Quality Standards for each pollutant. Copper has an upper concentration limit of 7 ug/L and Zinc has an upper limit of 50 ug/L. Based on the plots, the Copper concentrations in Eno River appear consistent between 2019 and 2020. However, the plot shows a significant decrease in Zinc concentrations between 2019 and 2020.

```{r echo=FALSE}
##ANOVA for metals
Zinc_anova <- aov(Eno_River_parameters_stations$Zinc ~ Eno_River_parameters_stations$Year)
summary(Zinc_anova)
 
Copper_anova <- aov(Eno_River_parameters_stations$Copper ~ Eno_River_parameters_stations$Year)
summary(Copper_anova) 
```

Metals: To determine whether there is a statistically significant change in Copper and Zinc concentrations from 2019 to 2020. A linear regression was performed for 2019 and 2020 Copper and Zinc data. Based on the p-value, the change in Copper from 2019 to 2020 was found not to be statistically significant. However, the difference in Zinc concentrations from 2019 to 2020 was determined to be statistically significant with a p-value < 0.01.    

```{r echo=FALSE}
#Temp and Sky condition
Temp_Sky.plot <- ggplot(subset(Eno_River_parameters_clean, Parameter == "Temperature"), aes(x = Sky.Condition, y = Final_Value, fill = Station.Name)) +
  geom_boxplot() +
  scale_fill_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  labs(y = "Temperature (Celcius)", x = "Sky Condition") +
  theme+
  theme(panel.grid.major = element_blank())
show(Temp_Sky.plot)

#Temp and Sky condition over time
Temp_Sky_ts.plot <- ggplot(subset(Eno_River_parameters_clean, Parameter == "Temperature"), aes(x = Month, y = Final_Value, color = Sky.Condition, shape = Station.Name)) +
  geom_point() +
  scale_color_viridis_d(option = "magma", begin = 0.8, end = 0.2, name = "Sky Condition") +
  #scale_x_date(date_breaks = "1 months", date_labels = "%b %Y") +
  labs(y = "Temperature (Celcius)", x = "Date") +
  theme+
  theme(panel.grid.major = element_blank(),
        axis.text.x = element_text(hjust = 1)) +
  facet_wrap(~Year, nrow = 2)
show(Temp_Sky_ts.plot)
```

Based on the boxplots, there does not appear to be significant difference in measured temperature based on the observed Sky Condition. The variation across the various Sky Conditions is not consistent across the different monitoring stations. However, there is a noticeable difference in temperature across stations which may give information about the monitoring location.

Across the year, the there is not a noticeable trend between the Sky Condition and the temperature. The temperature seems to be more seasonally affected than by the Sky Condition. Finer scale data, such as daily or hourly measurements would likely be needed in order to determine the impact of Sky Condition on Temperature. 


```{r echo=FALSE}
#turbidity and Recent Rain
Turbidity_Rain.plot <- ggplot(subset(Eno_River_parameters_clean, Parameter == "Turbidity"), aes(x = Rain.in.Last.24.Hours, y = Final_Value, fill = Station.Name)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.05), alpha = 0.7) +
  scale_fill_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  labs(y = "Turbidity (NTU)", x = "Rain in Last 24 Hours?") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow=1, scales = "free") +
show(Turbidity_Rain.plot)

Rain.plot <- ggplot(subset(Eno_River_parameters_clean,Parameter == "Turbidity"), aes(x = Year, y = Final_Value, fill = Rain.in.Last.24.Hours)) +
  geom_bar(width = 1, stat = "identity")
show(Rain.plot)
```
There is a noticeable difference in the turbidity concentrations and occurrence of rain within 24 hours This trend is less noticeable in the 2019, but the 2020 data shows a clear disparity between recent rain and the turbidity readings.

```{r echo=FALSE}
##ANOVA

#comparing turbidity and rain in last 24 hours for 2019 (all sites combined, combined sites because we have different numnber of sites between 2019 and 2020 so want to make the comparison similar)
d.2019 <- Eno_River_parameters_stations %>%
  filter(Year == 2019)

aov <- aov(d.2019$Turbidity ~ d.2019$Rain.in.Last.24.Hours)
summary(aov)

#comparing turbidity and rain in last 24 hours for 2020
d.2020 <- Eno_River_parameters_split %>%
  filter(Year == 2020)

aov <- aov(d.2020$Turbidity ~ d.2020$Rain.in.Last.24.Hours)
summary(aov)


#compare turbidity between 2019 and 2020 
aov <- aov(Eno_River_parameters_stations$Turbidity ~ Eno_River_parameters_stations$Year)
summary(aov)
```

Further analysis shows that the difference in Turbidity from rain within the last 24 hours was only statistically significant in 2020, with a p-value < 0.01. There are almost four times more observations in 2020 than 2019 which may have impacted the ANOVA results for 2019 since there were only 10 observations. Additionally, an ANOVA of the Turbidity across years, showed that the difference in values from 2019 to 2020 was not statistically significant. 

```{r echo=FALSE}
#TSS & Turbidity
Turbidity_TSS.plot <- ggplot(subset(Eno_River_parameters_clean,Parameter %in% c("Turbidity", "Total Suspended Solids")), aes(x = Month, y = Final_Value, color = Parameter, shape= Station.Name)) +
  geom_point() +
  geom_line()+
  scale_fill_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  labs(y = "Turbidity (NTU)", x = "Rain in Last 24 Hours?") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow = 2)
show(Turbidity_TSS.plot)

#TSS and Turbidity Linear Reg
TSS.regression <- lm(Eno_River_parameters_stations$`Total Suspended Solids`~Eno_River_parameters_stations$Turbidity)
summary(TSS.regression)


```

The plot shows that TSS and Turbidity readings are closely linked and follow the same trend across stations and across years. These parameters have similar peaks. There is some variation across stations, but the pattern remains the same across. The results of the ANOVA show that the difference between turbidity and TSS is not statistically significant. Additionally, the r-squared value tells us that 97% of the variance in Total Suspended Solids can be explained by variance in Turbidity.


```{r echo=FALSE}
# Temp & DO 
Temp_DO.plot <- ggplot(Eno_River_parameters_stations, aes(x = Temperature, y = `Dissolved Oxygen`, color = Station.Name)) +
  geom_point() +
  geom_line()+
  scale_fill_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  labs(y = "Dissolved Oxygen", x = "Temperature (Celcius)") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow = 2)
show(Temp_DO.plot)


#Temp & Fec Col
Temp_FC.plot <- ggplot(Eno_River_parameters_stations, aes(x = Temperature, y = `Fecal Coliform`, color = Station.Name)) +
  geom_point() +
  geom_line()+
  scale_fill_viridis_d(begin = 0.7, end = 0.2, name = "Sampling Station") +
  labs(y = "Fecal Coliform (cfu/100mL)", x = "Temperature (Celcius)") +
  theme+
  theme(panel.grid.major = element_blank())+
  facet_wrap(~Year, nrow = 2)
show(Temp_FC.plot)
```

##DO & Temp
There is a clear relationship with Temperature and DO. The DO decreases linearly as temperature increases. This is true for all monitoring stations and both years.

##FC & Temp
For the 2019 data there is not a clear relationship between Fecal Coliform concentration and temperature. However, in 2020 there is a peak in the Fecal Coliform concentration at the higher temperatures. It can not be concluded that there is a relationship between these parameters, but further investigation should occur. 

```{r echo=FALSE}
#Temp and DO linear regression
DO.regression <- lm(Eno_River_parameters_stations$`Dissolved Oxygen`~Eno_River_parameters_stations$Temperature)
summary(DO.regression)

#DO Yearly
DO_Year.regression <- lm(Eno_River_parameters_stations$`Dissolved Oxygen`~Eno_River_parameters_stations$Year)
summary(DO_Year.regression)

#Temp and Fecal Coliform linear regression
FC.regression <- lm(log(Eno_River_parameters_stations$`Fecal Coliform`)~Eno_River_parameters_stations$Temperature)
summary(FC.regression)

#FecalColiform Correlation Test
FC_Year.regression <- lm(log(Eno_River_parameters_stations$`Fecal Coliform`)~Eno_River_parameters_stations$Year)
summary(FC_Year.regression)

```
Based on the linear regression performed on temperature and Dissolved Oxygen there is significant negative correlation between temperature and Dissolved Oxygen concentrations. The correlation is nearly negative 1 describing the relationship of increased temperature on Dissolved Oxygen concentrations. Additionally, the r-squared value is nearly 0.85 which means that 85% of the variance in Dissolved Oxygen can be accounted for by temperature. However, there was no statistical difference in Dissolved Oxygen from 2019 to 2020. 

However, the variance in Fecal Coliform cannot be explained by temperature. The r-squared value is very low, 0.13, which shows that there is not a statistically significant relationship between these parameters. Additionally, there is not a statistically significant difference between 2019 and 2020 value of Fecal Coliform.  

```{r}
library(corrplot)
Paramcor <- cor(Eno_River_parameters_stations[15:16])
corrplot(Paramcor, method= "ellipse")
```

